\section{Introduction}
\label{sec:Intro}

In this paper we show how functional programming features (for example, the use of lazy evaluation) can be exploited in the context of modelling massive dynamically changing networks. Such networks are increasingly ubiquitous and provide important value-added to modern commerce and society. With growing size, however, comes additional computational overheads, and it is important to identify low-complexity, preferably sub-linear, algorithms for such heavily utilised tasks as path-finding, re-routing and dynamic workload distribution.

Our contribution in this paper is to provide a novel functional programming data structure for network representation that allows the programming of suitable low-complexity algorithms, and we provide experimental confirmation of our performance claims. As explained in more detao; below, we work with collections of spanning trees (one for each connected component of the network), and in this context our operations show the following complexities (Figure)

\begin{center}
\begin{tabular}{||l | l | c||} 
\hline
Function & Description & Complexity \\ %[0.5ex] 
\hline\hline
\texttt{linkTree} 
  & 
  & $O(m\log n)$ \\
\texttt{cutTree} 
  &
  & $O(\log n)$ \\
\hline
\texttt{conn} 
  & 
  & $O(\log n)$ \\ 
\texttt{link} 
  & 
  & $O(m\log n)$ \\
\texttt{cut} 
  & 
  & $O(\log n)$ \\
\hline
\texttt{reroot} 
  &  
  & $O(\log n)$ \\
\texttt{root} 
  & 
  & $O(1)$ \\
\hline   
\end{tabular}
\caption{Amortised time complexity for various dynamic network-editing operations in terms of their effect on the associated forest of spanning trees. In each case, $n$ gives the number of vertices in the first (or only) tree operated upon; for those functions taking two trees as input, $m$ i the number of verticies in the second tree. The result for \link assumes that $m \leq n$ (if not, we can swap the order of the arguments before applying the algorithm).
\normalsize 
\end{center}



give first work-optimal bounds for managing dynamic trees for purely functional data structures. We do this not only for extending the application of finger trees, we offer a solution for dynamic connectivity when facing persistent data structures as well as providing a simple interface to the user. The \dyntset exposes two types, \texttt{Tree} and \texttt{Forest}, to the user. The following is a compilation of \tcb{OUR} the functions, types and times we have implemented in Haskell, on top of that done by Hinze and Paterson \cite{FTs}:

\begin{center}
\small
 \begin{tabular}{||l | l | c||} 
 \hline
 Function & Type & Time complexity \\ %[0.5ex] 
 \hline\hline
 \texttt{linkTree} & \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{Tree} & $O(m\log n)$ \\
 \texttt{cutTree} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{(Tree,Tree)} & $O(\log n)$ \\
 \hline
 \texttt{conn} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Bool} & $O(\log n)$ \\ 
 \texttt{link} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Forest} & $O(m\log n)$ \\
 \texttt{cut} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Forest} & $O(\log n)$ \\
 \hline
 \texttt{reroot} & \texttt{Tree} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} & $O(\log n)$ \\
 \texttt{root} & \texttt{Tree} $\to$ \texttt{Maybe(Vertex)} & $O(1)$ \\
 \hline   
\end{tabular}
where $n$ and $m$ are the corresponding sizes of the trees involved; when $linking$ is performed, then the constraint $m \leq n$ should taken into account
\end{center}
\normalsize 


In general, the number of vertices in such a network can change from one moment to the next, and as new edges are introduced and existing one deleted, the underlying topology of the network will change, with connected components merging and separating as time passes. The problem we address is one of the simplest we can ask about a network, but at the same time one of the most fundamental, namely: \emph{given two vertices, $u$ and $v$, do they belong to the same component}? The answer, of course, will depend on when we ask the question, because the system we're considering is \emph{dynamic}; new connections can be created between vertices and old ones can be deleted.

If we are to answer this question efficiently, it follows that we need to use a dynamic data type to store information about the network, its vertices, and edges. Our goal is to ensure that updates to the network will result in only local changes to the data representation; we do not want to recompute the entire representation simply because one edge has been added or removed. Our approach is first to represent connected components not as graphs but as spanning trees, and then to use a novel approach to representing these trees dyanmically. [discussion: While adding a vertex to a component and adding it to the spanning tree are equivalent tasks, breaking an edge in the spanning tree corresponds to breaking multiple edges in the underlying component.]



Given our use of spanning trees to represent connected components, this question becomes: \emph{do $u$ and $v$ belong to the same tree}? Notice that adding new links within an already connected component does not affect its existing spanning trees; although new spanning trees may become possible, no existing spanning tree loses that status. On the other hand, if we connect vertices from distinct components, joining those nodes in the associated spanning trees automatically creates a new spanning tree for the larger component created by adding the link. (The relationship between deleting an edge in a component vs. deleting it in its spanning tree is more complex--see Sect.~\ref{sec:discussion}.)  We accordingly assume that the forest can change dynamically in response to repeated applications of two basic operations: \link and \cut. 
\begin{itemize}
\item
  if vertices $u$ and $v$ belong to different trees, $\link(u,v)$ adds the edge $(u,v)$ to the forest, thereby causing the trees containing $u$ and $v$ to be joined together to form a new, larger, tree; if the vertices already belong to the same tree, the operation has no effect. Formulated in this way, the function \link, can be denoted as
\item
  if vertices $u$ and $v$ are connected by an edge, $\cut(u,v)$ removes that edge, thereby causing the tree that contains it to be split into two smaller trees; if the vertices are not connected, the operation has no effect. 
\end{itemize}

\madd{WORKING HERE}

The problem has been studied as \textit{dynamic trees} under different variants according to the application, for instance, Sleator and Tarjan defined the \textit{link-cut} tree for solving network flow problems \cite{DS-DynTs}, Henzinger and King provided \textit{Euler tour} (ET) trees for speeding up dynamic connectivity on graphs \cite{Rand-DynGs-Algos} or Frederickson describing the \textit{topology} tree for maintaining the minimum spanning tree \cite{DSs-Online-Upd-MSTs}. The former and latter trees are based on techniques called \textit{path-decomposition} and \textit{tree contraction} respectively. In this paper we are interested on Euler tour trees, a technique called \textit{linearisation}, since the original tree (of any degree) is flattened and handled it as a sequence, in other words, turning a non-linear structure into a linear one.

All the above techniques have been implemented and studied for ephemeral data structures, performing $O(\log n)$ per operation. Functional data structures, on the other hand, ease the reasoning and implementation for the cases when preserving history is needed as in self-adjusting computation \cite{DynamizingAlgos} or as in version control \cite{CVS-Demaine}. To overcome the lack of pointers, researchers have devised data structures to represent efficient sequences, most notably, \textit{finger trees} \cite{FTs}. This structure performs well, allowing updates and queries in logarithmic time. We introduce the \emph{dynamic trees through Euler-tours}, a variant of the Hinze's and Paterson finger trees, or \dyntset for short. Like the finger tree, the \dyntset edit sequences in logarithmic time while providing dynamic tree operations. The key insight is to make $O(1)$ steps to perform \textit{connected}, \textit{link}, and \textit{cut} in logarithmic time. In Section \ref{sec:TechDes} we show this for \textit{connected} and \textit{cut}, whereas \textit{link} takes $O(m \log n)$. In Section \ref{sec:Eval} we show that, in practice, different results can be obtained, such $O(1)$ per operation as long as \link is interleaved with \cut.
 \tcb{NO FURTHER explanation about the $O(1)$ performance}

\madd{MOVED THIS TO TOP -- NEEDS A BIT IF REWRITING}

The contribution of this paper is to give first work-optimal bounds for managing dynamic trees for purely functional data structures. We do this not only for extending the application of finger trees, we offer a solution for dynamic connectivity when facing persistent data structures as well as providing a simple interface to the user. The \dyntset exposes two types, \texttt{Tree} and \texttt{Forest}, to the user. The following is a compilation of \tcb{OUR} the functions, types and times we have implemented in Haskell, on top of that done by Hinze and Paterson \cite{FTs}:

\begin{center}
\small
 \begin{tabular}{||l | l | c||} 
 \hline
 Function & Type & Time complexity \\ %[0.5ex] 
 \hline\hline
 \texttt{linkTree} & \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{Tree} & $O(m\log n)$ \\
 \texttt{cutTree} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} $\to$ \texttt{(Tree,Tree)} & $O(\log n)$ \\
 \hline
 \texttt{conn} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Bool} & $O(\log n)$ \\ 
 \texttt{link} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Forest} & $O(m\log n)$ \\
 \texttt{cut} & \texttt{Vertex} $\to$ \texttt{Vertex} $\to$ \texttt{Forest} $\to$ \texttt{Forest} & $O(\log n)$ \\
 \hline
 \texttt{reroot} & \texttt{Tree} $\to$ \texttt{Vertex} $\to$ \texttt{Tree} & $O(\log n)$ \\
 \texttt{root} & \texttt{Tree} $\to$ \texttt{Maybe(Vertex)} & $O(1)$ \\
 \hline   
\end{tabular}
where $n$ and $m$ are the corresponding sizes of the trees involved; when $linking$ is performed, then the constraint $m \leq n$ should taken into account
\end{center}
\normalsize 

The first block (first two rows) contains the functions to perform dynamic tree operations. The following block (rows 4, 5 and 6) lists the functions that compute an unbounded, although finite, sequence of dynamic tree operations over the same forest $F$. Finally, last two rows are the core functions, apart from the ones provided by Hinze and Paterson work \cite{FTs}. 
%, to perform dynamic tree operations. \tcb{PERHAPS explain the above table widely or paraphrasing ???}

In the next section, we give an overview of the finger tree functions and their performance. These are the foundations for our functions altogether with some of the functions from the Hackage library \texttt{Data.Set} \footnote{https://hackage.haskell.org/package/containers-0.5.10.2/docs/Data-Set.html} which mimic the set-like operations over our functions, such as testing membership. 

%In Section~\ref{sec:Example}, we present an in-depth example using the operations mentioned above. In particular, we depict how the \dyntset manages the trees as sequences under the dynamic setting, showing how the \dyntset structure is constructed immutably. 
In Section \ref{sec:TechDes}, we describe our implementation for performing dynamic tree operations with purely data structures, the \dyntset. The following is the address where our source code is hosted and publicly available:
\begin{center}
\url{https://github.com/jcsaenzcarrasco/ETdynTs}
\end{center}

We evaluate \dyntset empirically in Section~\ref{sec:Eval}. In concrete, we plot the results of benchmarking the functions in the first two sections of the table above. Our evaluation demonstrates that \dyntset performs according the function definitions with a constant factor between \conn and \link and \cut. 

%Section~\ref{sec:Discuss} discuss the main reasons we took into account when designing ETFT, with some suggestions for improving the internal vertex annotations. 
In Section~\ref{sec:RelWrk} we suggest potential alternatives for designing dynamic data structures to solve the problem described in the present document. 

Finally, we conclude in Section~\ref{sec:Concl} with a summary of our work.


\tcr{Presentationally I'm missing a better description of the background material. The paper recalls the definition of a monoid, but does not recall the definition of an Euler tour nor give enough detail on finger trees to read as an independent document.} 

