
\documentclass[runningheads,a4paper]{llncs}

\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx} 

% HASKELL SYNTAX HIGHLIGHTING - MUST BE LOADED BEFORE xcolor and listings IF THEY'RE ALSO BEING USED
\usepackage{mpshaskell}
\newcommand{\code}[1]{\haskell{#1}}
\newCodeWord{tree,list,original,reversed,x,xs,f,f',f'',y,ys,tour_x,tour_y,ex,ey,h,h',h'',e,e1,e2}
\newFunction{empty,singleton,member,reroot,splice,error,select,link,fromList',cut,path,guard,father}
\newSymbol{\insertL}{\lhd}
\newSymbol{\insertR}{\rhd}
% END OF HASKELL SYNTAX HIGHLIGHTING

% ADDED BY MIKE
\newcommand{\madd}[1]{\textcolor{Purple}{#1}}
\newcommand{\mdel}[1]{\textcolor{Yellow}{#1}}

\usepackage{xspace}
\newcommand{\MATHSF}[1]{\ensuremath{\mathsf{#1}}\xspace}
\newcommand{\link}{\MATHSF{link}}
\newcommand{\cut}{\MATHSF{cut}}
\renewcommand{\root}{\MATHSF{root}}
\newcommand{\reroot}{\MATHSF{reroot}}
\newcommand{\connected}{\MATHSF{connected}}
% END OF MIKE'S ADDITIONS


\usepackage{xcolor}

\usepackage{url}
\urldef{\mailjm}\path|{jcsaenzcarrasco1, m.stannett}@sheffield.ac.uk|    
\newcommand{\keywords}[1]{\par\addvspace\baselineskip
\noindent\keywordname\enspace\ignorespaces#1}

\newcommand{\tcr} [1]{\textcolor{red}{#1}}
\newcommand{\tcb} [1]{\emph{\textcolor{blue}{#1}}}


\begin{document}

\mainmatter  % start of an individual contribution

\title{Dynamic Trees through Euler Tours\\ \large{A purely functional programming approach}}

\titlerunning{Dynamic trees through Euler-tours, functionally}

\author{Juan Carlos Saenz-Carrasco%
\thanks{Postgraduate research student, supported by the Consejo Nacional de Ciencia y Tecnolog\'{i}a, CONACYT (the Mexican National Council for Science and Technology) under grant 411550, scholar 580617 and CVU 214885}%
\and Mike Stannett}

\authorrunning{Dynamic trees through Euler-tours, functionally}

\institute{Department of Computer Science,\\
Regent Court, 211 Portobello,\\
Sheffield S1 4DP, United Kingdom\\
\mailjm\\
\url{https://www.sheffield.ac.uk/}}


\toctitle{ETFT: a dynamic tree}
\tocauthor{J. Saenz, M. Stannett}
\maketitle


\begin{abstract}
We present purely-functional data structures for managing dynamic updates and queries. By doing so, we combine the structure of the Hinze's and Paterson finger tree with an efficient binary search tree to handle Euler tours as sequences: like its imperative counterpart, our proposal supports $O(\log n)$ amortised per operation, where $n$ is the number of the vertices in the tree. While existing data structures provide these time bounds, none do so to provide the three basic operations on dynamic trees: \link, \cut and \connected altogether within the same functional structure. However, in the nearly twenty years since, no one has shown that these functions work efficient under persistent data structures. \tcb{Some efforts have been done, for instance \url{chahine.moreau@gmail.com} at \url{https://github.com/k0ral/euler-tour-tree}. However there no proofs about correctness or efficiency. Also the part corresponding to forests is missing}
To understand how the algorithms perform in practice we have also implemented them in the Haskell functional programming language. Our implementations are not as fast as the best of these on the ephemeral and imperative realm but comparable by the same order of magnitude.
\keywords{Functional programming, persistent data structures, Haskell, \tcb{Experimental results}}
\end{abstract}


\input{./Intro/Intro}


\tcr{---------- P O S I T I V E    C O M M E N T S ------------------}


\tcr{This paper introduces Euler Tour Finger Trees (ETFTs), a persistent, purely-functional tree and forest data structure that supports O(log n) complexity for all operations, including link and cut. It gives an implementation in Haskell.}

\tcr{At a high-level, the tree is represented as an Euler Tour (a list of vertex pairs describing a path through the tree), which in turn is represented by a finger tree, thereby allowing O(log n) concatenation and splitting. That allows manipulating Euler tours efficiently, and building interesting tree operations easily.}

\tcr{Most of the implementation in this paper thus builds on existing implementations of finger trees and sets. Yet the combination seems novel, and the complexity bounds of the ETFT operations follow almost trivially, which is nice.
The paper presents a functional data structure for maintaining a dynamic forest under link and cut operations. It realizes this in O(log n) time complexity per operation (amortized), using an Euler tour tree representation implemented using a combination of finger trees and sets. }

\tcr{The paper first recalls dynamic trees and their operations, and then monoids, sets implemented with balanced search trees, Euler-tour trees and finger trees. It then gives (graphical) examples of how the sets and finger trees fit together to realize dynamic trees, it recalls graphically the root, reroot, split, concatenation, and view operations of finger trees, and then walks through the implementation of root, reroot, connected, link, and cut.
Finally the paper contains experiments illustrating that the implementation meets the claimed complexity bounds and discusses related and future work.
This is an interesting piece of work.}
%\end{comment}


\input{./Prelim/Prelim}

%\input{./Example/Example}

\input{./TechDes/TechDes} 

\input{./Eval/Eval} 

\input{./RelWrk/RelWrk} 

\input{./Concl/Concl} 

\bibliography{./Refs/refs}
\bibliographystyle{splncs03}

%\begin{thebibliography}{4}
%\input{./Refs/refs.bib}
%\end{thebibliography} 

\end{document}
